### 2020-10-05 manuscript/manuscript.rmd
small changes to the manuscript based on Nicolas's comments

p

@@

**Fig. 3.**  Summary of phylogenomic results from previous studies and reanalysis conducted in this study. Each point represents a phylogenetic analysis with a support of bootstrap values less than `r bootstrap_threshold`% or a posterior probability less than `r posterior_prob_threshold`% are considered as unresolved. The clades are organized with increased outgroup sampling higher on the vertical axis, and the models are organized in general by increasing complexity in terms of numbers of parameters to the right on the horizontal axis. (A) Analyses transcribed from the literature (related to Supplementary Table 2). (B) New phylogenomic analyses conducted in this study (related to Supplementary Table 3).

@@

Our new analyses show that, with restricted outgroup sampling, analyses of the same matrices with two different means of accommodating site heterogeneity in equilibrium frequencies (C60 in IQ-TREE and CAT in PhyloBayes) yield different results. This indicates that the traditional framing of the problem, that accommodating rate heterogeneity leads to support for Porifera-sister, is not correct. Instead, there is something about PhyloBayes CAT analyses specifically that leads to support for Porifera-sister.

@@

Taken together, this suggests that unconstrained CAT analyses are overparameterized. While models that accommodate site heterogeneity in equilibrium frequency are important in phylogenetic analyses, and tools such as PhyloBayes and IQ-TREE that implement such models are critical to phylogenetic questions considered here, the present analyses suggest that there are problems with unconstrained CAT analyses that are prone to create a large number of categories for a small fraction of sites. In the context of the present question, this means we should be skeptical of Porifera-sister not only because it is obtained under such a narrow range of analysis conditions, but because the very large number of categories needed to recover this result are adding a large number of parameters that explain a small fraction of the data.

@@

Philippe2009 =
  analyses_sensitive %>%
  filter(matrix == "Philippe2009_only_choanozoa" )

@@

Whelan2017_full_only_choanozoa =
  analyses_sensitive %>%
  filter(matrix == "Whelan2017_full_only_choanozoa")

@@

ggarrange(p1, p2, p3, p4,  ncol  = 2, nrow = 2, labels = c("A", "B", "C", "D"), common.legend = TRUE)

@@

```
**Fig. 4.**  Sensitive analyses with representative data matrices analyzed by different number of equilibrium frequency categories (nCAT) in PhyloBayes. Statistical support values (posterior probabilities) were obtained from three data matrices using the site-heterogeneous Poisson+CAT model with different categories. (A) Phlippe2009_Choanozoa; (B) Philippe2009_Holozoa; (C) Whelan2017_full_Choanozoa; (D) Whelan2017_strict. Statistical support for Ctenophora-sister and Porifera-sister is indicated in orange and green, respectively. Support values from the sensitive analyses are shown in Supplementary Table 6.

@@

Overcoming the challenges associated with this difficult and important question goes beyond models. In particular, there are many organisms that are highly relevant to this problem, in particular better sampling of outgroups, ctenophores, and sponges, for which no genome or transcriptome data are available [@King:2017ie]. In particular, there are very few chromosome-level genome assemblies for animals outside of Bilateria. There will be important advances in genome sequencing in coming years that will have a major impact on this problem. This is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and reduce the possibility of contamination and address occupancy issues in transcripomic data. We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting difficult phylogenetic problem, and also to develop methods and tools that will help with other difficult phylogenetic problems.

@@

ggarrange(p1, p2, p3, p4, p5,  ncol  = 3, nrow = 2, labels = c("A", "B", "C", "D" , "E"), common.legend = TRUE)

ggsave("figures/Extended_Data_Fig3_raw.pdf", width=10, height=8)
```

**Extended Data Fig. 3.** Comparison of the tree topologies under different substitutional models in Whelan2017_strict data matrix. (A). WAG+C60. (B). Poisson+nCAT60. (C). Poisson+nCAT90. (D). Poisson+CAT. (E). GTR+CAT.

@@

ggarrange(p1, p2, p3, p4, p5,  ncol  = 3, nrow = 2, labels = c("A", "B", "C", "D" , "E"), common.legend = TRUE)

ggsave("figures/Extended_Data_Fig4_raw.pdf", width=10, height=8)
```
**Extended Data Fig. 4.** Comparison of the tree topologies under different substitutional models in Philippe2009_Choanimalia data matrix. (A). WAG+C60. (B). Poisson+nCAT60. (C). Poisson+nCAT150. (D). Poisson+CAT. (E). GTR+CAT.

>>>>

@@

![](figures/Figure3_published.png)
**Fig. 3.**  Summary of phylogenomic results from previous studies and reanalysis conducted in this study. Each point represents a phylogenetic analysis with a support of bootstrap values less than `r bootstrap_threshold`% or a posterior probability less than `r posterior_prob_threshold`% are considered as unresolved. The clades are organized with increased outgroup sampling higher on the vertical axis, and the models are organized in general by increasing complexity in terms of numbers of parameters to the right on the horizontal axis. (A) Analyses transcribed from the literature (related to Supplementary Table 2). (B) New phylogenomic analyses conducted in this study (related to Supplementary Table 3).

@@

Our new analyses show that, with restricted outgroup sampling, analyses of the same matrices with two different means of accommodating site heterogeneity in equilibrium frequencies (C60 in IQ-TREE and CAT in PhyloBayes) yield different results. This indicates that the traditional framing of the problem, that accommodating rate heterogeneity leads to support for Porifera-sister, is not correct. Instead, there is something about PhyloBayes CAT analyses specifically that leads to support for Porifera-sister. Moreover, we found that the branch lengths of Ctenophora taxa are generally longer than other animals with both site-homogeneous and site-heterogeneous models in many data matrices (*e.g.* Extended Data Figs. 3, 4).

@@

Taken together, although further analysis are warranted, this suggests that unconstrained CAT analyses might be overparameterized. While models that accommodate site heterogeneity in equilibrium frequency are important in phylogenetic analyses, and tools such as PhyloBayes and IQ-TREE that implement such models are critical to phylogenetic questions considered here, the present analyses suggest that there are problems with unconstrained CAT analyses that are prone to create a large number of categories for a small fraction of sites. In the context of the present question, this means we should be skeptical of Porifera-sister not only because it is obtained under such a narrow range of analysis conditions, but because the very large number of categories needed to recover this result are adding a large number of parameters that explain a small fraction of the data.

@@

Philippe2009 =
  analyses_sensitive %>%
  filter(model != "GTR+CAT60") %>%
  filter(matrix == "Philippe2009_only_choanozoa" )

@@

Whelan2017_full_only_choanozoa =
  analyses_sensitive %>%
  filter(model != "GTR+CAT60") %>%
  filter(matrix == "Whelan2017_full_only_choanozoa")

@@

p=ggarrange(p1, p2, p3, p4,  ncol  = 2, nrow = 2, labels = c("A", "B", "C", "D"), common.legend = TRUE)

@@

```
![](figures/Figure4_manuscript.png)
**Fig. 4.**  Sensitive analyses with representative data matrices analyzed by different number of equilibrium frequency categories (nCAT) in PhyloBayes. Statistical support values (posterior probabilities) were obtained from three data matrices using the site-heterogeneous Poisson+CAT model with different categories. (A) Phlippe2009_Choanozoa; (B) Philippe2009_Holozoa; (C) Whelan2017_full_Choanozoa; (D) Whelan2017_strict. Statistical support for Ctenophora-sister and Porifera-sister is indicated in orange and green, respectively. Support values from the sensitive analyses are shown in Supplementary Table 6.

@@

Overcoming the challenges associated with this difficult and important question goes beyond models. In particular, there are many organisms that are highly relevant to this problem, in particular better sampling of outgroups, ctenophores, and sponges, for which no genome or transcriptome data are available [@King:2017ie]. In particular, there are very few chromosome-level genome assemblies for animals outside of Bilateria. There will be important advances in genome sequencing in coming years that will have a major impact on this problem. This is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and reduce the possibility of contamination and address occupancy issues in transcriptomic data. We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting difficult phylogenetic problem, and also to develop methods and tools that will help with other difficult phylogenetic problems.

@@

p=ggarrange(p1, p2, p3, p4, p5,  ncol  = 3, nrow = 2, labels = c("A", "B", "C", "D" , "E"), common.legend = TRUE)

ggsave("figures/Extended_Data_Fig3_raw.pdf", width=10, height=8)
```
![](figures/Extended_Data_Fig3_manuscript.png)
**Extended Data Fig. 3.** Comparison of the tree topologies under different substitutional models in Whelan2017_strict data matrix. (A). WAG+C60. (B). Poisson+nCAT60. (C). Poisson+nCAT90. (D). Poisson+CAT. (E). GTR+CAT.

@@

p= ggarrange(p1, p2, p3, p4, p5,  ncol  = 3, nrow = 2, labels = c("A", "B", "C", "D" , "E"), common.legend = TRUE)

ggsave("figures/Extended_Data_Fig4_raw.pdf", width=10, height=8)
```
![](figures/Extended_Data_Fig4_manuscript.png)
**Extended Data Fig. 4.** Comparison of the tree topologies under different substitutional models in Philippe2009_Choanimalia data matrix. (A). WAG+C60. (B). Poisson+nCAT60. (C). Poisson+nCAT150. (D). Poisson+CAT. (E). GTR+CAT.
